\documentclass[runningheads]{llncs}

% ---------------------------------------------------------------
% Include basic ECCV package
 
% TODO REVIEW: Insert your submission number below by replacing '*****'
% TODO FINAL: Comment out the following line for the camera-ready version
\usepackage[review,year=2026,ID=*****]{eccv}
% TODO FINAL: Un-comment the following line for the camera-ready version
%\usepackage{eccv}

% OPTIONAL: Un-comment the following line for a version which is easier to read
% on small portrait-orientation screens (e.g., mobile phones, or beside other windows)
%\usepackage[mobile]{eccv}


% ---------------------------------------------------------------
% Other packages

% Commonly used abbreviations (\eg, \ie, \etc, \cf, \etal, etc.)
\usepackage{eccvabbrv}

% Include other packages here, before hyperref.
\usepackage{graphicx}
\usepackage{booktabs}

% The "axessiblity" package can be found at: https://ctan.org/pkg/axessibility?lang=en
\usepackage[accsupp]{axessibility}  % Improves PDF readability for those with disabilities.


% ---------------------------------------------------------------
% Hyperref package

% It is strongly recommended to use hyperref, especially for the review version.
% Please disable hyperref *only* if you encounter grave issues.
% hyperref with option pagebackref eases the reviewers' job, but should be disabled for the final version.
%
% If you comment hyperref and then uncomment it, you should delete
% main.aux before re-running LaTeX.
% (Or just hit 'q' on the first LaTeX run, let it finish, and you
%  should be clear).

% TODO FINAL: Comment out the following line for the camera-ready version
%\usepackage[pagebackref,breaklinks,colorlinks,citecolor=eccvblue]{hyperref}
% TODO FINAL: Un-comment the following line for the camera-ready version
\usepackage{hyperref}

% Support for ORCID icon
\usepackage{orcidlink}


\begin{document}

% ---------------------------------------------------------------
% TODO REVIEW: Replace with your title
\title{Recovering Metric Scale in Pretrained Image-to-3D Generators Without Backbone Retraining}

% TODO REVIEW: If the paper title is too long for the running head, you can set
% an abbreviated paper title here. If not, comment out.
\titlerunning{Metric Scale Recovery for Image-to-3D Generators}

% TODO FINAL: Replace with your author list. 
% Include the authors' OCRID for the camera-ready version, if at all possible.
\author{First Author\inst{1}\orcidlink{0000-1111-2222-3333} \and
Second Author\inst{2,3}\orcidlink{1111-2222-3333-4444} \and
Third Author\inst{3}\orcidlink{2222--3333-4444-5555}}

% TODO FINAL: Replace with an abbreviated list of authors.
\authorrunning{F.~Author et al.}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.

% TODO FINAL: Replace with your institution list.
\institute{Princeton University, Princeton NJ 08544, USA \and
Springer Heidelberg, Tiergartenstr.~17, 69121 Heidelberg, Germany
\email{lncs@springer.com}\\
\url{http://www.springer.com/gp/computer-science/lncs} \and
ABC Institute, Rupert-Karls-University Heidelberg, Heidelberg, Germany\\
\email{\{abc,lncs\}@uni-heidelberg.de}}

\maketitle


\begin{abstract}
  The abstract should concisely summarize the contents of the paper. 
  While there is no fixed length restriction for the abstract, it is recommended to limit your abstract to approximately 150 words.
  Please include keywords as in the example below. 
  This is required for papers in LNCS proceedings.
  \keywords{First keyword \and Second keyword \and Third keyword}
\end{abstract}


\section{Introduction}
\label{sec:intro}

Image-to-3D generative models have reached a level of visual fidelity where a single photograph
is sufficient to produce a detailed mesh or Gaussian splat of an object with plausible geometry
and texture. Foundations such as SAM 3D Objects~\cite{sam3d} and TRELLIS~\cite{trellis} train
on millions of synthetic and real objects, learning rich priors over 3D shape and surface
appearance. Yet these models share a fundamental blind spot: their outputs live in a canonical
unit cube, $[-0.5, 0.5]^3$, with no connection to physical units. A reconstructed bowl and a
reconstructed car are indistinguishable in scale.

This missing dimension --- metric scale --- is not a cosmetic limitation. Robotic manipulation
requires knowing whether a grasped object is five centimetres or fifty. Augmented reality demands
that a reconstructed mug placed on a physical table displaces a physically plausible volume.
Scene reconstruction from multiple objects depends on consistent metric embedding. Without
absolute scale, a 3D generative model is an artist's tool, not a spatial computing primitive.

\paragraph{Why metric scale is hard.}
Recovering metric scale from a single image is fundamentally under-constrained without a prior
on object size. The standard workarounds each carry a cost. \emph{Category-mean priors} assume
the object is an average specimen of its class --- a heuristic that incurs $\sim$10.9\% mean
absolute percentage error (MAPE) on NOCS-Real275~\cite{nocs} even when the category is known.
\emph{Separate monocular metric depth models}~\cite{metric3d,zoedepth} can estimate depth in
physical units but require an additional inference pass, introduce a second set of parameters,
and still require a geometric bridge between pixel-depth and object volume.
\emph{Backbone retraining} --- modifying the 3D generator itself to reason about physical scale
--- demands compute budgets exceeding 100K GPU-hours for models of this class, which would also
destroy the generalisation properties built into the pretrained weights.

\paragraph{The hidden anchor.}
We observe that modern image-to-3D pipelines based on monocular depth estimation already contain
the metric signal needed for scale recovery --- it is simply being discarded. SAM 3D Objects uses
MoGe~\cite{moge} to lift the input image into a metric pointmap: a dense field of $(x, y, z)$
coordinates in metres aligned to the camera frame. From this pointmap the pipeline extracts a
global scale statistic (\texttt{pointmap\_scale}) and a vertical offset
(\texttt{pointmap\_shift\_z}) to normalise input conditioning --- and then never uses the
original metric values again. The object's true physical extent, encoded in those pointmap
statistics and in the spatial distribution of pointmap tokens, is present at inference time
but plays no role in the output.

\paragraph{Our approach.}
We propose a lightweight recipe for injecting metric awareness into a frozen image-to-3D
generator without retraining any backbone component. A compact \emph{MetricScaleHead} fuses
MoGe pointmap statistics with pooled sparse-structure (SS) latent features into a
1024-dimensional scale token. This token is injected into the spatial latent (SLAT)
flow-matching decoder via cross-attention --- the only part of the generative model that is
fine-tuned ($\sim$100M of more than one billion total parameters), with an fp32 upcast recipe
that prevents the bf16 attention overflow we encountered in early experiments. Separately, a
differentiable \emph{aspect-ratio loss} on the SS decoder's soft voxel occupancy supervises
object shape proportions directly, closing the gap between the metric head's per-axis scale
predictions and the mesh's actual bounding-box dimensions.

\paragraph{Results.}
Trained on the real-world NOCS-Real275 subset of OmniNOCS~\cite{omninocs} and evaluated on a
held-out scene split, our method achieves \textbf{1.23\% MAPE} (0.20\,cm mean absolute error)
--- an order of magnitude below the category-prior baseline and competitive with specialised
monocular metric depth methods that require a separate inference pass. Extended to
mixed-source training on Objectron~\cite{objectron} and ARKitScenes~\cite{arkitscenes}, the
model predicts metric scale across object categories, viewpoints, and indoor environments
without sacrificing per-category accuracy on the original evaluation set.

\paragraph{Contributions.} We contribute:
\begin{itemize}
  \item The observation that metric pointmap statistics are a latent anchor already present in
        image-to-3D pipelines and amenable to lightweight supervision.
  \item A practical recipe --- MetricScaleHead, scale-token cross-attention injection, and soft
        aspect-ratio supervision --- that adds metric scale to a frozen 3D generator with
        $\sim$170M trainable parameters and no backbone modification.
  \item A training stability protocol (fp32 cross-attention upcast, linear lr warmup) that
        makes fine-tuning large sparse flow-matching decoders viable without NaN instabilities.
  \item Empirical validation across three real-world object datasets demonstrating sub-2\%
        MAPE with generalisation to unseen object categories.
\end{itemize}

\section{Related Work}
\label{sec:related}

\paragraph{Image-to-3D generative models.}
Large-scale image-to-3D systems have converged on a two-stage latent flow-matching design. TRELLIS~\cite{trellis}
introduces the Structured 3D Latent (SLAT) representation, a sparse voxel grid carrying DINOv2 visual features
extracted from multi-view renders; a rectified-flow transformer decodes SLAT tokens conditioned on a single image,
with separate VAE heads producing Gaussian splats or textured meshes via FlexiCubes. TRELLIS v2~\cite{trellis2}
replaces the multiview-supervised features with native O-Voxel representations --- a ``field-free'' sparse structure
that handles open, non-manifold, and fully-enclosed surfaces --- achieving a $16\times$ spatial compression rate
and adding explicit PBR material support across four billion parameters. SAM 3D~\cite{sam3d} extends the SLAT
backbone to natural images with significant clutter and occlusion: a 1.2B-parameter Mixture-of-Transformers
geometry model jointly estimates coarse shape and object layout, followed by a 600M-parameter sparse latent
refinement model, all trained through a multi-stage synthetic-to-real flywheel aligned to human preference.
Despite their differences, all three systems output objects normalised to a unit cube with no connection to
physical dimensions. Our work keeps these frozen backbones intact and recovers the discarded metric scale
via a lightweight head trained on real-world object datasets.

\paragraph{Monocular geometry and metric depth estimation.}
Recovering scene geometry in physical units from a single image has seen rapid recent progress.
ZoeDepth~\cite{zoedepth} fine-tunes a relative-depth backbone with per-domain metric bins, enabling
zero-shot metric depth across indoor and outdoor scenes.
Metric3D~\cite{metric3d} achieves scale-aware depth through a canonical camera normalization module that
reconciles training on datasets with heterogeneous camera intrinsics.
MoGe~\cite{moge} reframes the problem as \emph{affine-invariant} pointmap prediction: it estimates a dense
$(x,y,z)$ field using a robust optimal-alignment loss, achieving state-of-the-art relative geometry accuracy
while deliberately leaving global scale and shift unconstrained. MoGe-2~\cite{moge2} extends this by
appending a lightweight MLP head on the ViT CLS token to predict a global metric scale factor, enabling
accurate metric geometry without sacrificing relative geometry quality. MoGe is integrated as a front-end in
both SAM 3D and the SAM 3D Objects pipeline we build on: the pipeline computes alignment statistics
(\texttt{pointmap\_scale}, \texttt{pointmap\_shift\_z}) to map MoGe's output into a unit-cube conditioning
space --- and then discards these normalization factors before the generative decoder runs. Our key
observation is that these statistics are correlated with the object's true physical extent, and routing them
directly into a supervised scale head recovers metric dimensions without any additional inference pass.

\paragraph{Category-level size and pose estimation.}
NOCS~\cite{nocs} defines the normalised object coordinate space framework for recovering 6-DoF pose and
metric bounding-box dimensions for known object categories; the accompanying NOCS-Real275 benchmark
(6 categories, 7k images) is our primary evaluation set.
OmniNOCS~\cite{omninocs} substantially extends the NOCS paradigm, unifying canonical 3D bounding-box
annotations across ten data sources --- including NOCS-Real275, Objectron~\cite{objectron},
ARKitScenes~\cite{arkitscenes}, KITTI, nuScenes, and Hypersim --- to produce a dataset of 97 object
categories and 380k images, twenty times the class coverage and 200 times the instance count of
prior NOCS datasets. The accompanying NOCSformer model predicts NOCS coordinates, instance masks,
and metric 3D boxes from 2D detections via a class-agnostic ViT backbone, generalising to unseen datasets
without class-specific heads. All category-level methods share a common limitation: metric scale recovery
is implicitly conditioned on category-level size priors encoded during training, which incurs a baseline MAPE
of $\sim$10.9\% on NOCS-Real275 even with ground-truth category labels. Our method uses no category
labels at inference time --- scale is inferred from pointmap normalization statistics and sparse latent
features that are category-agnostic --- and is trained on the same OmniNOCS sources.

\paragraph{Scale recovery in frozen generative models.}
Adapter-style fine-tuning has proven effective for injecting new capabilities into large frozen models in both
language~\cite{} and vision domains~\cite{}, but the objective has typically been semantic (e.g., style, content
control) rather than metric. In 3D, methods such as Zero123, InstantMesh, and CRM recover geometry in
normalised space; recovering metric scale from their outputs requires an independent calibration step.
To our knowledge, no prior work injects a metric supervision signal into the latent decoder of a large image-to-3D
generative model while preserving its frozen backbone weights. Our MetricScaleHead and cross-attention injection
recipe fill this gap.


\section{Method}
\label{sec:method}

\subsection{Overview}

We build on the SAM 3D Objects image-to-3D pipeline~\cite{sam3d}, which chains a monocular geometry
front-end (MoGe~\cite{moge}), a sparse structure (SS) generator that produces a coarse 3D occupancy
field, and a spatial latent (SLAT) flow-matching decoder that refines geometry and texture into a
Gaussian splat or mesh. All backbone parameters are frozen throughout our procedure.

Our method inserts three new components. (i)~A \emph{MetricScaleHead} consumes MoGe pointmap
statistics and SS latent features and produces a compact \emph{scale token}. (ii)~The scale token
is appended to the SLAT decoder's conditioning sequence so that its cross-attention layers can
attend to it; those layers are the only backbone weights that receive gradient updates during
fine-tuning. (iii)~A \emph{MetricScaleDecoder} reads the SLAT refined latent together with
the scale token and regresses log-space object dimensions $[\log W, \log H, \log D]$ in metres.
Separately, an \emph{aspect-ratio loss} on the SS decoder's soft voxel occupancy provides a
complementary shape-proportion signal that propagates back through the SS decoder's trainable
weights (the SS generator/encoder stays frozen).

\subsection{MetricScaleHead}
\label{sec:head}

The MetricScaleHead converts available metric signals into a single 1024-dimensional conditioning
token compatible with the SLAT backbone's context space.

\paragraph{Inputs.}
The head takes four inputs that are already computed by the standard pipeline at inference time,
requiring no additional forward passes:
\begin{enumerate}
  \item \textbf{Pooled SS shape latent.} The SS generator produces a $4096 \times 8$ sparse
        latent; global average pooling yields an 8-dimensional summary of object morphology.
  \item \textbf{SS scale features.} The SS generator's Mixture-of-Transformers (MoT) decoder
        outputs a dedicated \texttt{scale} token --- a 3-dimensional vector of log-SSI (scale-,
        shift-, and intensity-invariant) object extents supervised during pretraining with a
        loss weight of 0.1. These three values carry full-image DINOv2 and pointmap context
        already absorbed by the SS condition embedder.
  \item \textbf{$\log(\texttt{pointmap\_scale})$.} MoGe outputs an affine-invariant
        pointmap~\cite{moge}; the pipeline aligns it to the unit cube and records the
        alignment scale factor. The logarithm of this factor is a scalar correlated with
        the object's physical depth extent in the camera frame.
  \item \textbf{$\texttt{pointmap\_shift\_z}$.} The vertical (Z-axis) shift of the alignment
        in metres encodes the object's absolute distance from the camera, providing a complementary
        metric anchor to the scale factor.
\end{enumerate}
The 13-dimensional concatenation $[8 + 3 + 1 + 1]$ is fed through a three-layer MLP with
SiLU activations (widths $64 \to 128 \to 1024$), and a learned modality embedding
$\boldsymbol{e}_\text{metric} \in \mathbb{R}^{1024}$ is added to the output:
\begin{equation}
  \mathbf{s} = \mathrm{MLP}\bigl([\bar{\mathbf{z}}_\text{SS};\, \mathbf{f}_\text{scale};\,
    \log s_\text{pm};\, \Delta_z]\bigr) + \boldsymbol{e}_\text{metric} \;,
  \label{eq:head}
\end{equation}
where $\bar{\mathbf{z}}_\text{SS}$ is the pooled SS latent, $\mathbf{f}_\text{scale}$ the
SS scale token, $s_\text{pm}$ the pointmap alignment scale, and $\Delta_z$ the vertical shift.
The modality embedding ensures that SLAT cross-attention can distinguish the injected scale
token from the DINOv2 image and mask tokens that normally constitute the conditioning sequence.

\subsection{Scale-Token Injection via Cross-Attention}
\label{sec:injection}

The SLAT backbone is a sparse rectified-flow transformer with 24 blocks, each containing a
self-attention layer followed by a cross-attention layer that reads from the condition embedder's
output. We inject the scale token $\mathbf{s} \in \mathbb{R}^{1 \times 1024}$ by appending it
to the cross-attention context sequence without modifying any backbone code:
\begin{equation}
  \mathbf{C}' = \bigl[\mathbf{C}_\text{base};\, \mathbf{s}\bigr] \;,
  \label{eq:inject}
\end{equation}
where $\mathbf{C}_\text{base}$ contains the standard DINOv2 image and mask tokens.
This is implemented as a transparent proxy (\texttt{\_ScaleAugmentedEmbedderProxy}) that wraps
the original condition embedder, so the SLAT generator's forward pass is unchanged.

We selectively unfreeze the cross-attention sublayers (\texttt{cross\_attn} and \texttt{norm2})
across all 24 blocks, totalling $\sim$100\,M of the more than one billion total SLAT parameters.
All self-attention weights and feed-forward blocks remain frozen.

\paragraph{fp32 cross-attention upcast.}
The SLAT backbone runs in bfloat16. Preliminary experiments found that cross-attention keys
and values computed from the unbounded scale token caused \textsc{NaN} overflow in bf16.
We wrap each unfrozen \texttt{cross\_attn} and \texttt{norm2} module with an fp32 shim that
casts inputs up before the forward pass and casts outputs back to bfloat16 afterwards. This
keeps the training numerically stable without any change to the model architecture.

\paragraph{CFG-style scale-token dropout.}
During training, with probability $p=0.1$ the scale token injected into the SLAT context
sequence is replaced with a zero vector (while the unmodified token is still passed to
MetricScaleDecoder). This mirrors classifier-free guidance (CFG) dropout from TRELLIS~\cite{trellis}
and encourages the cross-attention layers to function gracefully when the scale token
is absent, reducing overfitting to its magnitude.

\subsection{MetricScaleDecoder}
\label{sec:decoder}

After the SLAT flow-matching forward pass, the refined sparse latent provides per-voxel features
$\mathbf{Z} \in \mathbb{R}^{N_v \times 8}$ (where $N_v$ is the number of active voxels). A
scatter mean pools these features per batch element, yielding an 8-dimensional geometry
summary. This is concatenated with a 16-dimensional linear projection of the scale token:
\begin{equation}
  \hat{\mathbf{d}} = \mathrm{MLP}\bigl([\overline{\mathbf{Z}};\, \mathbf{W}_s \mathbf{s}]\bigr)
    \in \mathbb{R}^3 \;,
  \label{eq:decoder}
\end{equation}
where $\mathbf{W}_s \in \mathbb{R}^{16 \times 1024}$. The three output logits are interpreted as
$\hat{\mathbf{d}} = [\log \hat{W}, \log \hat{H}, \log \hat{D}]$ and exponentiated at inference
time to produce metric dimensions in metres. The log-space parameterisation keeps predictions
positive by construction and normalises the target distribution across object scales spanning
two orders of magnitude.

\subsection{Aspect-Ratio Supervision via Soft Voxel Occupancy}
\label{sec:ratio}

The SS decoder outputs a $64^3$ soft occupancy volume (pre-sigmoid logits) as a byproduct of
generating the coarse structure. We exploit this to provide a complementary shape-proportion
signal that does not require a metric anchor --- only the \emph{relative} extents of the predicted
shape. For each axis $a \in \{x, y, z\}$ we compute the marginal occupancy distribution
$p_a$ by summing over the other two axes, then derive a differentiable per-axis extent:
\begin{equation}
  \sigma_a = \sqrt{\mathrm{Var}_{p_a}[i_a]} \;,
  \label{eq:sigma}
\end{equation}
where $i_a$ are voxel indices along axis $a$. To handle the axis-permutation ambiguity (the
canonical voxel frame need not align with physical width/height/depth), we rank-sort
$[\sigma_x, \sigma_y, \sigma_z]$ and $[W, H, D]$ in descending order before comparing.
The aspect-ratio loss is a scale-invariant smooth-L1 on mean-centred log extents:
\begin{equation}
  \mathcal{L}_\text{AR} = \mathrm{SmoothL1}\!\left(
    \log \boldsymbol{\sigma}_\downarrow - \overline{\log \boldsymbol{\sigma}_\downarrow},\;
    \log \mathbf{d}_\downarrow - \overline{\log \mathbf{d}_\downarrow}
  \right) \;,
  \label{eq:ratio}
\end{equation}
where $\downarrow$ denotes rank-sorted descending. Subtracting the mean removes the global
scale, so the loss penalises only shape-proportion errors. To receive gradients, the SS
decoder's head weights are unfrozen; the SS generator and encoder remain frozen.

\subsection{Training Objective and Optimisation}
\label{sec:training}

The primary objective is a smooth-L1 loss on predicted vs.\ ground-truth log dimensions:
\begin{equation}
  \mathcal{L} = \mathrm{SmoothL1}(\hat{\mathbf{d}},\, \log \mathbf{d}_\text{GT})
    + \lambda_\text{AR}\,\mathcal{L}_\text{AR} \;,
  \label{eq:loss}
\end{equation}
with $\lambda_\text{AR} = 0.05$ in all experiments. The smooth-L1 kernel is less sensitive to
outlier objects than MSE while retaining a quadratic regime near zero for stable convergence.

\paragraph{Optimiser and learning rates.}
MetricScaleHead and MetricScaleDecoder are trained with Adam at $\text{lr} = 10^{-4}$, weight
decay $10^{-4}$. The unfrozen SLAT cross-attention weights use a separate parameter group
at $\text{lr} = 10^{-6}$, linearly warmed up from zero over 500 steps to prevent
catastrophic interference with the pretrained attention weights at the start of fine-tuning.
Gradient norms are clipped per parameter group using an adaptive percentile-based clipper.

\paragraph{Warm-start and dataset.}
All experiments warm-start from a checkpoint trained on the NOCS-Real275 subset of
OmniNOCS~\cite{omninocs}. Mixed-source training extends to Objectron~\cite{objectron} and
ARKitScenes~\cite{arkitscenes} with a per-source cap of 30,000 records and balanced
sampling (WeightedRandomSampler with weight $\propto 1/|\text{source}|$) to prevent any
single dataset from dominating the gradient signal.


\section{Experiments}
\label{sec:experiments}

TODO


\section{Conclusion}
\label{sec:conclusion}

TODO


\section*{Acknowledgements}
TODO

% ---- Bibliography ----
%
% BibTeX users should specify bibliography style 'splncs04'.
% References will then be sorted and formatted in the correct style.
%
\bibliographystyle{splncs04}
\bibliography{main}
\end{document}
